Probabilistic Catalogs for Crowded Stellar Fields

### 2013-03-01 paper/ms.tex
Lots of writing towards the end

Telescopes don't make catalogs \citep{2011EAS....45..351H}, they
measure the intensity field.  Viewed through the lens of probabilistic
inference, the goals of astronomy are to take the information in the
telescope-generated records of the intensity field and deliver it to
the quantities of astronomical interest with as little loss as
possible.  Insertion of a catalog-generation step in the inference
pipeline between the raw imaging and the final astrophysical analyses
is potentially lossy; the hard decisions of catalog making destroy
information.  Probability theory suggests that it may be less
lossy to pass forward not a catalog but a probabilisitic
description of all the catalogs that could be consistent with the
imaging---a posterior probability in (enormously large) space of possible catalogs.
Essentially, the creation of a catalog is an
attempt to answer the question,
``Given the image we have obtained, what objects are present in the field and
what are their properties?''.
This article represents an attempt at this ambitous goal in the specific
situation where the only objects in the field are stars or other point sources.

@@

The remainder of this paper is structured as follows. In Section~\ref{sec:bayes}
we give a brief overview of Bayesian inference. In Section~\ref{sec:model} we
discuss the model assumptions we make in our method. In Section~\ref{sec:mcmc}
we briefly discuss our
MCMC implementation. Section~\ref{sec:simulated_data} describes the tests we
carried out on simulated data, and a comparison with \sex~results is presented
in Section~\ref{sec:sex}. We conclude in Section~\ref{sec:conclusion}.

@@

For simplicity, we assume a uniform prior probability distribution for the
position of each star. This creates a strong preference for catalogs where
the stars are uniformly scattered across the image. Thus, this model is
appropriate for small patches of sky where the density of stars is approximately
uniform. In other scenarios, such as images of stellar clusters, it is possible
to parameterize the spatial distribution of the stars in a similar way to how
we have parameterized the luminosity function.

@@

Throughout this paper we will
assume the pixel-convolved PSF is a weighted mixture of two concentric
circular
Gaussians (a similar model is also used by \citet{1987PASP...99..191S}) with widths
$s_1$ and $s_2$:
\begin{eqnarray}
\mathcal{P}(x, y) = \frac{w}{2\pi s_1^2}\exp
\left[
-\frac{1}{2s_1^2}\left(x^2 + y^2\right)
\right]
+ \frac{1-w}{2\pi s_2^2}\exp
\left[-\frac{1}{2s_2^2}\left(x^2 + y^2\right)
\right].
\end{eqnarray}
The pixellated observed image is assumed to be generated from the mock
image (evaluated at the center of each pixel) plus noise:
\begin{eqnarray}
I_{ij} &=& \mathcal{M}(X_{ij}, Y_{ij}) + \epsilon_{ij}
\end{eqnarray}
where the errors $\{\epsilon_{ij}\}$ are independent and normally distributed.
The variance of each pixel is determined by the brightness of the sky and the
brightness of the mock image in that pixel:
\begin{eqnarray}
\epsilon_{ij} \sim \mathcal{N}(0, \sigma_0^2 + \eta \mathcal{M}(X_{ij}, Y_{ij})).
\end{eqnarray}
where $\sigma_0$ is a constant noise level and $\eta$ is a coefficient allowing
the noise to be higher in brighter regions of the image. This dependence of the
noise variances on the model intensity arises as a result of the Poissonian
nature of photon counts, but allows for the fact that a ``sky'' background may
have already been subtracted from the image.
This parameterisation has been used by \citet{2011MNRAS.412.2521B} and is an
alternative to the common practice of producing a ``variance map'' from the
image data that is then assumed to be known.

@@

From these samples, we can construct the
posterior expected true scene and other summaries. Summary images are shown in Figure~\ref{fig:summaries}. The residuals
provide a check on the validity of the model, and the posterior expected true
scene provides a useful visual guide to the uncertainties present in the catalogs.

\begin{figure}[ht!]
\begin{center}
\includegraphics[width=\textwidth]{Figures/summaries.eps}
\end{center}
\caption{Summary images for test case 2. The upper left panel shows the posterior
mean high-resolution scene. The upper right panel shows the posterior mean scene
when observed at the resolution of the data, and the bottom panels show the
standardized model residuals.
\label{fig:summaries}}
\end{figure}

As with test case 1, the PSF parameters $\{s_1, s_2, w\}$ and the noise parameters $\{\sigma_0, \eta\}$ were also inferred accurately with small uncertainties.

@@

In Figure~\ref{fig:luminosity_function}, we present the cumulative luminosity
function of the stars in the two fields, defined as the number of stars above
a given flux. The true cumulative luminosity function is plotted along with
posterior samples from the Bayesian method and catalogs produced by \sex.

\begin{figure}[ht!]
\begin{center}
\includegraphics[width=\textwidth]{Figures/luminosity_function.eps}
\end{center}
\caption{The cumulative luminosity functions produced by the Bayesian method
and
\sex, compared with the
actual cumulative LF. \sex~finds only a small fraction of the
stars actually present. This is to be expected in the high-$\sigma$ case as
\sex~then only searches for stars above a certain flux, but at the low-$\sigma$
end the number of stars is still drastically underestimated. Note that the true
numbers
of stars shown in these plots are $\sim$82\% of the values listed in
Figure~\ref{tab:truth} because here we only count those stars with positions
in $[-1, 1] \times [-1, 1]$.
\label{fig:luminosity_function}}
\end{figure}

For Test Case 2, we then took the catalog produced by \sex~for the $0.5\sigma$
threshold and fitted a broken power-law model to the list of fluxes in the catalog.
Such an approach might be used to study the luminosity function of a set of stars
in a field. The results of this inference are shown in Figure~\ref{fig:sex_inference}.
The inference spuriously rules out the true values of the parameters. This
occurs because the uncertainty in the stellar fluxes is not propagated through
to the inference of the luminosity function.

\begin{figure}[ht!]
\begin{center}
\includegraphics[width=\textwidth]{Figures/sex_inference.eps}
\end{center}
\caption{Inference of the luminosity function parameters using the $0.5\sigma$
catalog from \sex. Note that the inference appears to rule out regions around
the true input values. This occurs because the uncertainties in the existence
and fluxes of the faint stars are not taken into account. The true input values are plotted as filled squares.
\label{fig:sex_inference}}
\end{figure}

@@

In this paper we have developed and demonstrated a Bayesian approach to
making catalogs from astronomical images in the case where the image contains
only stars (or other point sources). The key idea is that instead of computing
a single catalog, the method creates a posterior probability distribution on
the space of possible catalogs that represents our state of knowledge about
the presence and properties of objects in the image. When this is done, the
uncertainties in the imaging are accurately propagated through to scientific
conclusions, for example about the luminosity function of the stars.
This approach was contrasted with a standard method of running \sex~and then
fitting a luminosity function to the fluxes in the catalog. This latter approach
was shown to produce incorrect inferences about the luminosity function.

We note that there are many limitations to the model presented in this paper.
In principle, our model should be a model of the physical state of the universe,
and not a simple model where the only stellar properties are a 2-D position and
a flux. Another limitation is that we have not considered multi-epoch or multi-band
imaging. In the former case, PSF variations and stellar motions may be relevant
\citep{lang}, and in the latter, a model for the spectral energy distributions
of the stars will need to be considered.

@@

Throughout this paper, we have also assumed that the pixel-convolved PSF can
be adequately modeled using simple components and that there are no PSF
variations across the field. Relaxing this assumption provides a significant
challenge for the future.

>>>>

Telescopes don't make catalogs \citep{2011EAS....45..351H}, they
measure the intensity field.  Viewed through the lens of probabilistic
inference, the goals of astronomy are to take the information in the
telescope-generated records of the intensity field and deliver it to
the quantities of astronomical interest with as little loss as
possible.  Insertion of a catalog-generation step in the inference
pipeline between the raw imaging and the final astrophysical analyses
is potentially lossy; the hard decisions of catalog making destroy
information, at least in principle.  Probability theory suggests that it may be less
lossy to pass forward not a catalog but a probabilisitic
description of all the catalogs that could be consistent with the
imaging---a posterior probability in (enormously large) space of possible catalogs.
Essentially, the creation of a catalog is an
attempt to answer the question,
``Given the image we have obtained, what objects are present in the field and
what are their properties?''.
This article represents an attempt at this ambitous goal in the specific
situation where the only objects in the field are stars or other point sources.

@@

In this paper, we develop a Bayesian object detection model with the following
features: i) the number $N$ of objects in the image is an unknown parameter
to be inferred, ii) the objects that we expect to find are point sources such
as stars, and iii) the point-spread function is unknown (but a parametric model is
used) and must be inferred from
the data (but a single bright star may not be available to help with estimating it).
This paper is structured as follows. In Section~\ref{sec:bayes}
we give a brief overview of Bayesian inference. In Section~\ref{sec:model} we
discuss the model assumptions we make in our method. In Section~\ref{sec:mcmc}
we briefly discuss our
MCMC implementation. Section~\ref{sec:simulated_data} describes the tests we
carried out on simulated data, and a comparison with \sex~results is presented
in Section~\ref{sec:sex}. We conclude in Section~\ref{sec:conclusion}.

@@

For simplicity, we assume a uniform prior probability distribution for the
position of each star. The use of independent priors for the positions creates
a strong preference for catalogs where
the stars are uniformly scattered across the image. Thus, this model is
appropriate for small images where the density of stars is approximately
uniform across the imaeg. In other scenarios, such as images of stellar clusters, it is possible
to parameterize the spatial distribution of the stars in a similar way to how
we have parameterized the luminosity function, i.e. as a hierarchical model.

@@

Throughout this paper we will
assume the pixel-convolved PSF is a weighted mixture of two concentric
circular
Gaussians (a similar mixture model is also used by DAOPHOT, \citet{1987PASP...99..191S}) with widths
$s_1$ and $s_2$:
\begin{eqnarray}
\mathcal{P}(x, y) = \frac{w}{2\pi s_1^2}\exp
\left[
-\frac{1}{2s_1^2}\left(x^2 + y^2\right)
\right]
+ \frac{1-w}{2\pi s_2^2}\exp
\left[-\frac{1}{2s_2^2}\left(x^2 + y^2\right)
\right].
\end{eqnarray}
The pixellated observed image is assumed to be generated from the mock
image (evaluated at the center of each pixel) plus noise:
\begin{eqnarray}
I_{ij} &=& \mathcal{M}(X_{ij}, Y_{ij}) + \epsilon_{ij}
\end{eqnarray}
where the errors $\{\epsilon_{ij}\}$ are independent and normally distributed.
The variance of the normal distribution for each pixel is determined by the
brightness of the sky and the
brightness of the mock image in that pixel. This can be modelled by assuming the
following distribution:
\begin{eqnarray}
\epsilon_{ij} \sim \mathcal{N}(0, \sigma_0^2 + \eta \mathcal{M}(X_{ij}, Y_{ij})).
\end{eqnarray}
where $\sigma_0$ is a constant noise level and $\eta$ is a coefficient allowing
the noise to be higher in brighter regions of the image. This dependence of the
noise variances on the model intensity arises as a result of the Poissonian
nature of photon counts, but allows for the fact that a ``sky'' background may
have already been subtracted from the image in the reduction process.
This parameterisation has been used by \citet{2011MNRAS.412.2521B} and is an
alternative to the common practice of producing a ``variance map'' from the
image data that is then assumed to be known.

@@

From these samples, we can construct the
posterior expected true scene and other summaries. Summary images are shown in Figure~\ref{fig:summaries}. The residuals
provide a check on the validity of the model assumptions, and the posterior expected true
scene provides a useful visual guide to the uncertainties present in the catalogs. In this example,
the residuals show only noise because the simulated images were actually generated from the model.

\begin{figure}[ht!]
\begin{center}
\includegraphics[width=\textwidth]{Figures/summaries.eps}
\end{center}
\caption{Summary images for Test Case 2. The upper left panel shows the posterior
mean high-resolution scene. The upper right panel shows the posterior mean scene
when observed at the resolution of the data, and the bottom panels show the
standardized model residuals.
\label{fig:summaries}}
\end{figure}

The inferences on the number of stars $N$ and the luminosity function parameters
are shown in Figure~\ref{fig:results2}. The uncertainty about the luminosity
function parameters is still considerable despite the larger number of stars, as the fluxes of the fainter stars
are not well constrained by the data. The true values are still well within
the range of plausible values in the posterior distribution.
As with Test Case 1, the PSF parameters $\{s_1, s_2, w\}$ and the noise parameters $\{\sigma_0, \eta\}$ were also inferred accurately with small uncertainties.

@@

In Figure~\ref{fig:luminosity_function}, we present the cumulative luminosity
function (CLF) of the stars in the two fields, defined as the number of stars above
a given flux. The true CLF is plotted along with several
posterior samples from the Bayesian method and catalogs produced by \sex~for
various values of the detection and analysis thresholds.
We note that the true CLF is typical of the posterior samples, as expected.
For both test cases, the \sex~catalog is also consistent with the posterior
distribution at the bright end. However, the inferences from \sex~and the Bayesian
method differ at the faint end, with the former significantly underestimating
the number of faint stars.

This result may be attributable to the fact that the Bayesian method knows
about the existence and form of the luminosity function, even though it
does not know the values of the parameters. To test this, we ran the inference
on the data using an incorrect exponential distribution for the luminosity
function. The resulting CLF from this run did underestimate the number of low stars.
However, the
Bayesian evidence for the exponential model was significantly lower (by a factor of
approximately $10^7$) than
for the (correct) broken power-law model.

\begin{figure}[ht!]
\begin{center}
\includegraphics[width=\textwidth]{Figures/luminosity_function.eps}
\end{center}
\caption{The cumulative luminosity functions produced by the Bayesian method
(several posterior samples shown)
and
\sex~(for various values of the threshold parameters), compared with the
actual cumulative LF. Both methods correctly identify the fluxes at the bright
end, with some uncertainty due to overlapping sources. However, at the lower
end \sex~is unable to detect all of the stars whereas the true CLF is typical
of the posterior distribution.
\label{fig:luminosity_function}}
\end{figure}

@@

In this paper we have developed and demonstrated a Bayesian approach to
making catalogs from astronomical images in the case where the image contains
only stars (or other point sources). The key idea is that instead of computing
a single catalog, the method creates a posterior probability distribution on
the space of possible catalogs that represents our state of knowledge about
the presence and properties of objects in the image. When this is done, the
uncertainties in the imaging are accurately propagated through to scientific
conclusions, for example about the luminosity function of the stars.
This approach was contrasted with the results from the standard \sex~software.
For the bright sources the results were essentially consistent, however the
Bayesian approach was more successful at modelling the distribution of faint
stars. Of course, the Bayesian method is much more computationally intensive,
which is a significant issue in practice. To build a more complete picture of
when our approach is necessary in practice, it will need to be tested against
a wider variety of alternative methods.

We note that there are many limitations to the model presented in this paper,
some of which will be important to relax when applying it to real data.
In principle, our model should be a model of the physical state of the universe,
and not a simple model where the only stellar properties are a 2-D position and
a flux. Another limitation is that we have not considered multi-epoch or multi-band
imaging. In the former case, PSF variations and stellar motions may be relevant
\citep{lang}, and in the latter, a model for the spectral energy distributions
of the stars will need to be considered: essentially, the luminosity function
will need to be a probability distribution over more than one dimension.

@@

Throughout this paper, we have also assumed that the pixel-convolved PSF can
be adequately modeled using simple components and that there are no PSF
variations across the field. Relaxing these assumption provides a significant
challenge for the future.
